---
title: "Importance Sampling, performance"
author: "Viktor Bluhme Jeppesen and Malte Nikolajsen"
date: "29/9/2020"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE)
```


```{r import, include=FALSE}
source("assignments/2_unisamp/main.R")
library(profvis)
library(microbenchmark)
library(bench)
library(tidyverse)
library(patchwork)
```

First set basic parameters:

```{r}
n <- 100                  # Length of each random walk (same as n in assignment text)
S0 <- 30                  # Initial wealth (= 30 in assignment)
X_min <- -1.9; X_max <- 2 # Support of X ~ Unif(X_min, X_max)
theta <- -0.3             # Theta in Importance Sampling
```

The probability of ruin is approximately

```{r}
ruin_IS_rcpp(10^6, n, S0, X_min, X_max, theta)[10^6]
```

We first check that the R and Rcpp implementations return same values. For Monte Carlo:

```{r}
N_sim <- 1000
set.seed(100)
A <- ruin_MC(N_sim, n, S0, X_min, X_max)
set.seed(100)
B <- ruin_MC_rcpp(N_sim, n, S0, X_min, X_max)

all.equal(A, B) # True
```

For Importance Sampling:

```{r}
set.seed(100)
A <- ruin_IS(N_sim, n, S0, X_min, X_max, theta)
set.seed(100)
B <- ruin_IS_rcpp(N_sim, n, S0, X_min, X_max, theta)

all.equal(A, B) # True
```

For Monte Carlo, confidence bands are given by $\pm 1.96 * \sqrt{\hat p_N (1-\hat p_N) / N}$, where
\[
\hat p_N = \frac1N \sum_{i=1}^N h(X^{(i)}), \quad h(X_i) = 1_{(\exists k \leq n \mid S_k \leq 0)}(X^{(i)})
.\]
These are shown here for one sample path of the cummulative mean:

```{r}
N_sim <- 10^5
tibble(N = 1:N_sim, cummean = ruin_MC_rcpp(N_sim, n, S0, X_min, X_max)) %>% 
  filter(N %% 10 == 0) %>% 
  mutate(p = cummean) %>% 
  ggplot(aes(x = N, y = cummean)) +
  geom_ribbon(
    mapping = aes(
      ymin = pmax(0, cummean - 1.96 * sqrt(p*(1-p) / N)),
      ymax = pmin(1, cummean + 1.96 * sqrt(p*(1-p) / N))
    ), fill = "gray") +
  coord_cartesian(ylim = c(0, 0.005)) +
  geom_line()
```

## Benchmarking

For a total of $1000$ simulations, the benchmark numbers look like:
```{r}
N_sim <- 10^3
bench::mark(
  "MC R" = {ruin_MC(N_sim, n, S0, X_min, X_max); 1},
  "MC Rcpp" = {ruin_MC_rcpp(N_sim, n, S0, X_min, X_max); 1},
  "IS R" = {ruin_IS(N_sim, n, S0, X_min, X_max, theta); 1},
  "IS Rcpp" = {ruin_IS_rcpp(N_sim, n, S0, X_min, X_max, theta); 1},
)
```
We now benchmark for $10, 100, 1000, 10000$ simulations and plot the median times.

```{r}
pressed <- bench::press(
  N_sim = 10^(1:4),
  {
    bench::mark(
      "MC R" = {ruin_MC(N_sim, n, S0, X_min, X_max); 1},
      "MC Rcpp" = {ruin_MC_rcpp(N_sim, n, S0, X_min, X_max); 1},
      "IS R" = {ruin_IS(N_sim, n, S0, X_min, X_max, theta); 1},
      "IS Rcpp" = {ruin_IS_rcpp(N_sim, n, S0, X_min, X_max, theta); 1},
      iterations = 100
    )
  }
)
```

```{r}
p1 <- autoplot(pressed)
p2 <- pressed %>% 
  mutate(Method = attr(expression, "description")) %>% 
  ggplot(aes(x = N_sim, y = median, color = Method)) +
  geom_line() +
  geom_point() +
  scale_x_log10() +
  theme(legend.position = "none") +
  ggtitle("Log-log plot")

p3 <- pressed %>% 
  mutate(Method = attr(expression, "description")) %>% 
  ggplot(aes(x = N_sim, y = median, color = Method)) +
  geom_line() +
  geom_point() +
  scale_y_continuous()  +
  ggtitle("Linear axes")

p1 / 
(p2 | p3) +
  plot_layout(heights = c(1.5, 1))
```

Importance Sampling is much more efficient than pure Monte Carlo integration.

```{r}
set.seed(1000)

## Plot of sample paths for cumulative mean
k <- 10         # Number of sample paths simulated
N_sim <- 10^5   # Number of simulations per sample path

cummean <- tibble()
for (i in 1:k) {
  MC <- ruin_MC_rcpp(N_sim, n, S0, X_min, X_max)
  IS <- ruin_IS_rcpp(N_sim, n, S0, X_min, X_max, theta)
  cummean <- cummean %>% bind_rows(tibble(type = "MC", i = i, N = 1:N_sim, cummean = MC))
  cummean <- cummean %>% bind_rows(tibble(type = "IS", i = i, N = 1:N_sim, cummean = IS))
}
cummean %>% 
  filter(N %% 100 == 0) %>%
  ggplot(aes(x = N, y = cummean, col = as.factor(i))) +
  geom_line() +
  facet_wrap(~ type) +
  coord_cartesian(ylim = c(0.001, 0.0025)) +
  theme(legend.position = "none")

```


