---
title: "Bivariate Smoothing (document)"
author: "Viktor Bluhme Jeppesen and Malte Nikolajsen"
date: "9/8/2020"
output: html_document


header-includes:
    - \newcommand{\argmin}{\mathop{\mathrm{argmin}}\limits}
---

```{r setup, include=FALSE}
library(tidyverse)
library(splines)
library(latex2exp)
library(microbenchmark)
library(testthat)
library(cowplot)
set.seed(1)
knitr::opts_chunk$set(
	fig.align = "center",
	fig.pos = "h",
	cache = TRUE,
	cache.lazy = TRUE,
	dev.args = list(bg = "transparent"),
	out.width = "70%"
)
theme_set(theme_bw(base_size=12))

Nuuk <- read_table2("nuuk.dat.txt", 
                    col_names = c("Year", 1:12), 
                    na = "-999", 
                    skip = 1) %>% 
  gather(key = "Month", value = "Temperature", -Year, convert = TRUE) %>% 
  mutate(Temperature = Temperature / 10) %>% 
  filter(Year > 1866)
Nuuk_year <- group_by(Nuuk, Year) %>% 
  summarise(
    Temperature = mean(Temperature),
    Median = median(Temperature),
    High = max(Temperature), 
    Low = min(Temperature)
  )
n <- nrow(Nuuk_year)
x <- Nuuk_year$Year
y <- Nuuk_year$Temperature
df <- tibble(x, y)
pp <- ggplot(df, aes(x, y))
```

## Splines

Given points $\{(x_1,y_1), \dots, (x_n, y_n)\}$, we want to implement the smoother which solves
\[
\min_{f \in C^2}\quad L(f) = \sum_{i=1}^n\big(y_i - f(x_i)\big)^2 + \lambda \|f''\|_2^2
\]
All solutions belong to the $n$-dimensional vector space consisting of cubic splines with knots in $\mathbf x = (x_1, \dots, x_n)$, independently of $\mathbf y$. Thus all candidate solutions can be written
\[
f = \sum_{i=1}^n \beta_i \varphi_i
\]
for a basis $\varphi_1, \dots, \varphi_n$ of cubic splines, or more concisely $\mathbf f = \boldsymbol \Phi \beta$ for $\mathbf f = (f(x_1), \dots, f(x_n))$ and $\boldsymbol \Phi_{ij} = \varphi_j(x_i)$. We use this to rewrite the loss function as
\[
L(f) = (\mathbf y - \boldsymbol \Phi\beta)'(\mathbf y - \boldsymbol \Phi\beta) + \lambda\beta'\boldsymbol \Omega\beta
\]
where
\[
\boldsymbol \Omega_{ij} = \langle\varphi_i'', \varphi_j''\rangle = \int \varphi_i''\varphi_j''\, dm.
\]
The minimizer is given by
\[
\hat{\mathbf{f}} = \boldsymbol\Phi(\boldsymbol\Phi'\boldsymbol\Phi + \lambda\boldsymbol\Omega)^{-1}\boldsymbol\Phi'\mathbf y \equiv \mathbf S_\lambda\mathbf y.
\]
We must overcome two computational challenges, namely 1) how to evaluate $\boldsymbol\Omega$ and 2) how to efficiently calculate $\mathbf S_{\lambda}$.

First, we find a function to evaluate $\boldsymbol\Omega$. From CSwR, using Simpson's rule:

```{r penalty_matrix}
pen_mat <- function(inner_knots) {
  knots <- sort(c(rep(range(inner_knots), 3), inner_knots))
  d <- diff(inner_knots)  # The vector of knot differences; b - a 
  g_ab <- splineDesign(knots, inner_knots, derivs = 2) 
  knots_mid <- inner_knots[-length(inner_knots)] + d / 2
  g_ab_mid <- splineDesign(knots, knots_mid, derivs = 2)
  g_a <- g_ab[-nrow(g_ab), ]
  g_b <- g_ab[-1, ]
  (crossprod(d * g_a,  g_a) + 
      4 * crossprod(d * g_ab_mid, g_ab_mid) + 
      crossprod(d * g_b, g_b)) / 6 
}
```

Steps for efficient calculation of $\mathbf S_\lambda = \boldsymbol\Phi(\boldsymbol\Phi'\boldsymbol\Phi + \lambda\boldsymbol\Omega)^{-1}\boldsymbol\Phi'$

1. Single-value decomposition $\Phi = UDV'$.

2. Diagonalize the matrix $D^{-1}V'\Phi VD^{-1} = W\Gamma W'$.

3. Then 
\begin{equation}\label{eq:S_decomp}
\mathbf S_\lambda = \widetilde U(I + \lambda \Gamma)^{-1}\widetilde U'
\end{equation}
for $\widetilde U = UW$.

Step-by-step version for implementation:

- First, the coefficients $\hat\beta = \widetilde U'y$ are computed for expanding $y$ in the basis given by the columns of $\widetilde U$.

- Second, the $i$-th coefficient is shrunk towards $0$,
\[
\hat\beta_i(\lambda) = \frac{\hat\beta_i}{1+\lambda\gamma_i}.
\]

- Third, the smoothed values $\widetilde U\hat\beta_i(\lambda)$ are computed as an expansion using the shrunken coefficients.

We implement the decomposition.
```{r decomposition}
myDecomposition <- function(x, y, p) {
  inner_knots <- seq(min(x), max(x), length.out = p - 2)
  
  Phi <- splineDesign(c(rep(range(inner_knots), 3), inner_knots), x)
  Omega <- pen_mat(inner_knots)
  
  Phi_svd <- svd(Phi)
  Omega_tilde <- t(crossprod(Phi_svd$v, Omega %*% Phi_svd$v)) / Phi_svd$d
  Omega_tilde <- t(Omega_tilde) / Phi_svd$d
  Omega_tilde_svd <- svd(Omega_tilde)  
  U_tilde <- Phi_svd$u %*% Omega_tilde_svd$u
  
  structure(list(U_tilde = U_tilde, Gamma = Omega_tilde_svd$d), class = "customDecomposition")
}
```

We are to find the optimal tuning parameter $\hat \lambda$ using LOOCV

\[
\hat \lambda 
=
\argmin_{\lambda \in \mathbb R_+}
\text{LOOCV}(\lambda)
=
\argmin_{\lambda \in \mathbb R_+}
\sum_{i = 1, \ldots, n}
\left(
\frac{
y_i - \hat f^{-i}
}{
1 - S_{i,i}
}
\right)^2
\]

In writing the `r`-implementation, it should expect class `customDecomposition`. 

In an efficient implementation, we should not compute the smoother matrix $S_\lambda$, only its diagonal. 
Letting $\text{diag}(A) = (a_{1,1}, \ldots, a_{m,m})$ for any square matrix $A \in M_{m, m}(\mathbb R)$, we observe for $\widetilde U(I + \lambda \Gamma)^{-1}\widetilde U' \in M_{n,n}(\mathbb R)$

\[
\text{diag}(\widetilde U(I + \lambda \Gamma)^{-1}\widetilde U')
=
\left (
  \sum_{j = 1}^p \tilde u_{i, j}^2 / (1 + \lambda_j \Gamma_{j,j})
\right )_{i = 1, \ldots, n}
\]

This gives the following efficient implementation of LOOCV. 
```{r loocv}
myLOOCV <- function(lambda, decomp, y) {
  expect_s3_class(decomp, "customDecomposition")
  f_hat <- t(decomp$U_tilde) %*% y
  f_hat <- f_hat / (1 + lambda * decomp$Gamma)
  f_hat <- decomp$U_tilde %*% f_hat
  diag <- decomp$U_tilde^2 %*% as.matrix(1 / (1 + lambda * decomp$Gamma))
  sum((y - f_hat)^2 / (1 - diag)^2)
}
```
Alternatively, we could just compute the smoothing matrix $S_\lambda$.
```{r loocv inefficient}
myAltLOOCV <- function(lambda, decomp, y) {
  expect_s3_class(decomp, "customDecomposition")
  S <- t(decomp$U_tilde)
  S <- S / (1 + lambda * decomp$Gamma)
  S <- decomp$U_tilde %*% S
  diag <- diag(S)
  sum((y - S %*% y)^2 / (1 - diag)^2)
}
```
```{r loocv simulation data, cache = TRUE, echo = FALSE}
n <- 10000
x.sim <- 1:n
y.sim <- c(0, cumsum(rnorm(n - 1)))
```
```{r loocv compared, echo = FALSE}
wrapper.MyLoocv <- function(p, lambda) {
  decomp <- myDecomposition(x.sim, y.sim, p)
  myLOOCV(lambda, decomp, y.sim)
}
wrapper.MyAltLoocv <- function(p, lambda) {
  decomp <- myDecomposition(x.sim, y.sim, p)
  myAltLOOCV(lambda, decomp, y.sim)  
}
myBenchOnLoocv <- press(
  lambda <- c(50, 100, 150),
  p <- c(4000, 5000, 6000),
  {
    data <- wrapper.MyLoocv(p, lambda)
    mark(data(p, lambda))
  }
)
myBenchOnLoocv
```







```{r smoothing, message=FALSE}
mySmoothing <- function(
  ..., data, p = NULL, lambda = NULL, m = 300
) {
  ord <- order(data$x)
  x <- data$x[ord]
  y <- data$y[ord]
  nx <- length(x)
  if (is.null(p)) {
    p <- .nknots.smspl(nx)
  }
  decomp <- myDecomposition(x, y, p)
  if (is.null(lambda)) {
    lambda <- optimize(myLOOCV, c(0, m), decomp = decomp, y = y, tol = )$minimum
  }
  f_hat <- t(decomp$U_tilde) %*% y
  f_hat <- f_hat / (1 + lambda * decomp$Gamma)
  f_hat <- decomp$U_tilde %*% f_hat
  structure(
    list(x = x, y = f_hat), class = "mySmoothing" 
  )
}
predict.mySmoothing <- function(
  object, newdata, ...
) {
  approx(object$x, object$y, newdata$x)$y
}
splineSmoothingR <- smooth.spline(df$x, df$y, cv = TRUE)
```
```{r smoothing-plots, echo = FALSE, cache = TRUE, out.height = 10, out.width = 12}
plot1 <- (pp + 
    labs(x = "Year", y = "Mean temperature") +
    theme(legend.position = "bottom") +
    geom_point() +
    geom_smooth(method = "mySmoothing", se = FALSE, aes(colour = "mySmoothing-method")) + 
    geom_line(aes(x, y, colour = "smooth.spline-method"), tibble(x = splineSmoothingR$x, y = splineSmoothingR$y)) +
    scale_colour_manual("", values = c("blue", "red")))
temp.mySmooth <- mySmoothing(data = df)
df.plot2 <- tibble(
  x = df$x,
  y = temp.mySmooth$y - splineSmoothingR$y)
plot2 <- (pp +
    labs(x = "Year", y = "Mean temperature fitted difference") +
    theme(legend.position = "bottom") +
    geom_line(aes(x, y, colour = "Numeric difference"), tibble(x = df$x, y = temp.mySmooth$y - splineSmoothingR$y)) + 
    scale_colour_manual("", values = c("black")))
plot_grid(plot1, plot2, rel_widths = c(1.5, 1.5))
```









