---
title: "Bivariate Smoothing"
author: "Viktor Bluhme Jeppesen and Malte Nikolajsen"
date: "17 September 2020"
output: ioslides_presentation
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
```

## To-do

- Implement a smoothing spline smoother using LOOCV for selecting the tuning parameter $\lambda$.
  - Using B-spline basis functions and their derivatives from `splines::splineDesign`

- Test implementation and compare with `smooth.spline`.

- Test on real data and simulated data (benchmarking, profiling)

Use matrix decompositions to speed up LOOCV.

Link to CSwR chapter: [Bivariate smoothing](https://cswr.nrhstat.org/3-bivariate.html).

## Splines

Given points $\{(x_1,y_1), \dots, (x_n, y_n)\}$, we want to implement the smoother which solves
\[
\min_{f \in C^2}\quad L(f) = \sum_{i=1}^n\big(y_i - f(x_i)\big)^2 + \lambda \|f''\|_2^2
\]
All solutions belong to the $n$-dimensional vector space consisting of cubic splines with knots in $\mathbf x = (x_1, \dots, x_n)$, independently of $\mathbf y$. Thus all candidate solutions can be written
\[
f = \sum_{i=1}^n \beta_i \varphi_i
\]
for a basis $\varphi_1, \dots, \varphi_n$ of cubic splines.

## Splines

More concisely $\mathbf f = \boldsymbol \Phi \beta$ for $\mathbf f = (f(x_1), \dots, f(x_n))$ and $\boldsymbol \Phi_{ij} = \varphi_j(x_i)$. We use this to rewrite the loss function as
\[
L(f) = (\mathbf y - \boldsymbol \Phi\beta)'(\mathbf y - \boldsymbol \Phi\beta) + \lambda\beta'\boldsymbol \Omega\beta
\]
where
\[
\boldsymbol \Omega_{ij} = \langle\varphi_i'', \varphi_j''\rangle = \int \varphi_i''\varphi_j''\, dm.
\]
The minimizer is given by
\[
\hat{\mathbf{f}} = \boldsymbol\Phi(\boldsymbol\Phi'\boldsymbol\Phi + \lambda\boldsymbol\Omega)^{-1}\boldsymbol\Phi'\mathbf y \equiv \mathbf S_\lambda\mathbf y.
\]
We must overcome two computational challenges, namely 1) how to evaluate $\boldsymbol\Omega$ and 2) how to efficiently calculate $\mathbf S_{\lambda}$.

## Penalty matrix

First, we find a function to evaluate $\boldsymbol\Omega$. From CSwR, using Simpson's rule:

```{r}
pen_mat <- function(inner_knots) {
  knots <- sort(c(rep(range(inner_knots), 3), inner_knots))
  d <- diff(inner_knots)  # The vector of knot differences; b - a 
  g_ab <- splineDesign(knots, inner_knots, derivs = 2) 
  knots_mid <- inner_knots[-length(inner_knots)] + d / 2
  g_ab_mid <- splineDesign(knots, knots_mid, derivs = 2)
  g_a <- g_ab[-nrow(g_ab), ]
  g_b <- g_ab[-1, ]
  (crossprod(d * g_a,  g_a) + 
      4 * crossprod(d * g_ab_mid, g_ab_mid) + 
      crossprod(d * g_b, g_b)) / 6 
}
```
